---
title: "EDA gh.v2"
author: "Barbara E. Mottey"
date: "11/8/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(tidyverse)
library(ggmosaic)
library(GGally)
library(forcats)
```

## Load data

```{r}
gh_data <- readRDS("named_datagh.rds")

```

##birth weight and fuel
```{r}
gh.data <- gh_data
attach(gh.data)
summary(c_weight) # 28,699 NAs Min is 500 and max is 9000
table(fuel_bin) # 801 NAs

#(k<-gh.data %>% group_by(year) %>% summarise(observations= n()))
filter(gh_data, c_weight> 5500) %>% summarise(n=n())  # 34 obs are greater than 5500


#bw 5884, 1971 not weighed (coded 9996), dont know 543 NA 3
#2008 2903 2290 613 weighed 
## read in, filter by bw<=5500 and compare with what the FRQ file has.
```

# refine data
```{r}
# remove bw above 5500 and bw NA's 
gh.data <- gh.data %>% filter(c_weight <= 5500) #28746 total removed

#filter data from year 2000

filter(gh.data, year_cmc> 2000) %>% summarise(n=n()) # total of 4299 data 
gh.data <- gh.data %>% filter(year_cmc > 2000)

#save data
saveRDS(gh.data, "model_datagh.rds")
```




## birth weight distribution
```{r}
ggplot(gh.data, aes(c_weight))+geom_histogram(binwidth = 10.5, alpha = 1.0)+ ylim(0,800)+stat_bin(bins= 100, na.rm = T) + xlab("birth weight")
```



## birthweight vs year per region
```{r}
ggplot(gh.data)+ aes(as.factor(year_cmc), c_weight)+geom_point()+ facet_wrap(~region, labeller=(as_labeller(c("1" = "Western", "2"  = "Central", "3"= "Greater Accra", "4"= "Volta", "5"="Eastern", "6"= "Ashanti", "7"="Brong Ahafo", "8"= "Northern", "9"= "Upper East", "10"= "Upper West")))) +labs(x= "year", y = "birth weight", title = 'Scatter plot of birthweight per region')

ggplot(gh.data)+ aes(as.factor(year_cmc), c_weight)+geom_boxplot()+ facet_wrap(~region, labeller=(as_labeller(c("1" = "Western", "2"  = "Central", "3"= "Greater Accra", "4"= "Volta", "5"="Eastern", "6"= "Ashanti", "7"="Brong Ahafo", "8"= "Northern", "9"= "Upper East", "10"= "Upper West")))) +labs(x= "year", y = "birth weight", title = 'Boxplot of birthweight per region')


ggplot(gh.data, aes(c_weight))+geom_histogram(binwidth = 10.5, alpha = 1.0)+ ylim(0,100)+stat_bin(bins= 200, na.rm = T) +facet_wrap(~region, labeller=(as_labeller(c("1" = "Western", "2"  = "Central", "3"= "Greater Accra", "4"= "Volta", "5"="Eastern", "6"= "Ashanti", "7"="Brong Ahafo", "8"= "Northern", "9"= "Upper East", "10"= "Upper West")))) +labs(x= "birth weight", title = 'Histogram of birthweight per region')
```


## boxplots for fuel types
```{r}
gh.data$fuel_bin[gh.data$fuel_bin == 2]<- NA
gh.data$fuel_bin <- as.factor(gh.data$fuel_bin)

ggplot(na.omit(gh.data), aes(x= fuel_bin, y= c_weight))+geom_boxplot(outlier.shape = NA)+ ylim(0,6000) + ylab("birth weight") + xlab("fuel type")+
  scale_x_discrete(labels = c("0" = "Clean", "1"  = "Solid"))
```
Generally, people who use solid fuel tend to have babies with lower birthweights comparatively.


##per region
```{r}
ggplot(na.omit(gh.data), aes(x= fuel_bin, y= c_weight))+geom_boxplot(outlier.shape = NA)+ ylim(0,6000)+scale_x_discrete(labels = c("0" = "clean", "1"  = "solid"))+facet_wrap(~region, labeller=(as_labeller(c("1" = "Western", "2"  = "Central", "3"= "Greater Accra", "4"= "Volta", "5"="Eastern", "6"= "Ashanti", "7"="Brong Ahafo", "8"= "Northern", "9"= "Upper East", "10"= "Upper West")))) +labs(x= "fuel", y= "birth weight", title = 'Boxplot of fuel type used vs birthweight per region')
```

Stratified by region, there seems to be no difference in birthweights for babies with mothers that use solid or clean fuel.


###fuel type  vs wealth
```{r}
gh.data$residence<- as.factor(gh.data$residence)
gh.data$wealth <- as.factor(gh.data$wealth)

ggplot(na.omit(gh.data)) + geom_mosaic(aes(product(wealth, fuel_bin), fill = wealth)) + facet_wrap(~residence, labeller=(as_labeller(c("1" = "urban", "2" = "rural")))) + ylab("wealth") + xlab("fuel (clean/solid)")

ggplot(na.omit(gh.data)) + geom_mosaic(aes(product(residence, wealth), fill = residence)) +
  facet_wrap(~fuel_bin, labeller=(as_labeller(c("0" = "Clean fuel", "1"  = "Solid fuel")))) +labs(x= "wealth", y = "residence", title = 'fuel type (clean/solid) used based on residence type and wealth')
```
#Facet by Residential
Among urban residents, a higher proportion of those who use clean fuels are the rich people. About the 80% of those who use solid fuel are at least average people in terms of wealth.
However among the rural residents, a larger proportion (about 90%) use solid fuel. Most of those who use sold fuel in the rural setting are poor where as those who use clean fuel (about 98%) have at least average wealth.

#Facet by Fuel
The higher the wealth index, the more likely the residence is urban for the clean fuel users.

## Fuel vs wealth
```{r}
ggplot(na.omit(gh.data))+ geom_boxplot(aes(wealth, c_weight, fill = fuel_bin, alpha = fuel_bin))+
  labs(x = "wealth", y = "Birth weight", title = "birth weight vs wealth")+
  scale_x_discrete(labels = c("1"= "Poorest", "2"= "Poor", "3" = "Middle", "4" = "Rich", "5" = "Richest"))+
  scale_alpha_ordinal(name = "fuel type", labels = c("clean", "solid"))+
  scale_fill_discrete(name = "fuel type", labels = c("clean", "solid"))
```



##Fuel vs education
```{r}
gh.data$education <- as.factor(gh.data$education)

ggplot(na.omit(gh.data)) + geom_mosaic(aes(product(education, fuel_bin), fill = education)) + labs(x= "fuel type", y = "education", title = 'fuel type (clean/solid) used based on level of education')
```

Most of the women that use clean fuel have a secondary education; same applies to those who use solid fuel. About 50% of those who use solid fuel have at most primary education.


## fuel vs smoking
```{r}
gh.data$smoke <- as.factor(gh.data$smoke)

ggplot(na.omit(gh.data)) + geom_mosaic(aes(product(smoke, fuel_bin), fill = smoke)) + labs(x= "fuel type", y = "smoking status", title = 'fuel type (clean/solid) used based on smoking status')
```

Smoking status would not be a covariate. Most women do not smoke.


##education vs birth weight
```{r}

ggplot(na.omit(gh.data), aes(x= education, y= c_weight))+geom_boxplot(outlier.shape = NA)+ ylim(0,6000)+ ylab("birth weight")+scale_x_discrete(labels = c("0" = "none", "1"  = "primary", "2"="Secondary", "3"="tertiary"))
```

Those with no education or primary education seem to have babies with a lower mean weight.


##birth weight vs wealth
```{r}
ggplot(na.omit(gh.data), aes(x= wealth, y= c_weight))+geom_boxplot(outlier.shape = NA)+ ylim(0,6000)+scale_x_discrete(labels = c("1" = "poorest", "2"  = "poor", "3"="average", "4"="rich", "5"="richest" ))
```

The mean birth weight for each category seem to be lower for those who are poorest or poor.

## birth weight vs residence
```{r}
ggplot(na.omit(gh.data), aes(x= residence, y= c_weight))+geom_boxplot(outlier.shape = NA)+ ylim(0,6000)+scale_x_discrete(labels = c("1" = "urban", "2"  = "rural"))
```

Average Birth weights for those in rural settings seem to be quite lower than those in the urban setting.


## birth weight vs gender
```{r}
gh.data$gender <- as.factor(gh.data$gender)
ggplot(na.omit(gh.data), aes(x= gender, y= c_weight))+geom_boxplot(outlier.shape = NA)+ ylim(0,6000)+ labs(x= "gender", y= "birth weight")+scale_x_discrete(labels = c("1" = "male", "2"  = "female"))
```

boys seem to have a higher average birth weight than girls


