---
title: "Ghana birthweight: Exploratory"
author: "Barbara E. Mottey"
date: "10/12/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyverse)
library(ggmosaic)
library(GGally)
library(forcats)
```

## Load data

```{r}
dhs <- readRDS("new_dhs.rds")

gh_data<- dhs %>% select(-c("notes", "isoalpha2")) %>% filter(country.name=="Ghana")
```

##birth weight and fuel
```{r}
gh.data <- gh_data
attach(gh.data)
summary(c_weight) # 27,595 NAs Min is 800 and max is 9000
table(fuel_bin) # 801 NAs

```



## birth weight distribution
```{r}
ggplot(gh.data, aes(c_weight))+geom_histogram(binwidth = 10.5, alpha = 1.0)+ ylim(0,4000)+stat_bin(bins= 100, na.rm = T) + xlab("birth weight")
```


## boxplots for each country for birth weight
```{r}

ggplot(gh.data , aes(y= c_weight)) +
  geom_boxplot(fill = '#e5e5e5', outlier.shape = NA) +
  ylab("birth weight") 
```


## boxplots for fuel types
```{r}
gh.data$fuel_bin[gh.data$fuel_bin == 2]<- NA
gh.data$fuel_bin <- as.factor(gh.data$fuel_bin)

ggplot(na.omit(gh.data), aes(x= fuel_bin, y= c_weight))+geom_boxplot(outlier.shape = NA)+ ylim(0,6000) + ylab("birth weight") + xlab("fuel type")+
  scale_x_discrete(labels = c("0" = "Clean", "1"  = "Solid"))
```

People who use solid fuel tend to have lower birthweights comparatively.

###fuel type  vs wealth
```{r}
gh.data$residence<- as.factor(gh.data$residence)
gh.data$wealth <- as.factor(gh.data$wealth)

ggplot(na.omit(gh.data)) + geom_mosaic(aes(product(wealth, fuel_bin), fill = wealth)) + facet_wrap(~residence, labeller=(as_labeller(c("1" = "urban", "2" = "rural")))) + ylab("wealth") + xlab("fuel (clean/solid)")

ggplot(na.omit(gh.data)) + geom_mosaic(aes(product(residence, wealth), fill = residence)) +
  facet_wrap(~fuel_bin, labeller=(as_labeller(c("0" = "Clean fuel", "1"  = "Solid fuel")))) +labs(x= "wealth", y = "residence", title = 'fuel type (clean/solid) used based on residence type and wealth')
```
#Facet by Residential
Among urban residents, a higher proportion of those who use clean fuels are the rich people. About the 80% of those who use solid fuel are at least average people in terms of wealth.
However among the rural residents, a larger proportion (about 90%) use solid fuel. Most of those who use sold fuel in the rural setting are poor where as those who use clean fuel (about 98%) have at least average wealth.

#Facet by Fuel
The higher the wealth index, the more likely the residence is urban for the clean fuel users.


##Fuel vs education
```{r}
gh.data$education[gh.data$education %in% c(4:9)]<- NA
gh.data$education <- as.factor(gh.data$education)

ggplot(na.omit(gh.data)) + geom_mosaic(aes(product(education, fuel_bin), fill = education)) + labs(x= "fuel type", y = "education", title = 'fuel type (clean/solid) used based on level of education')
```

Most of the women that use clean fuel have a secondary education; same applies to those who use solid fuel. About 50% of those who use solid fuel have at most primary education.


## fuel vs smoking
```{r}
gh.data$smoke <- as.factor(gh.data$smoke)

ggplot(na.omit(gh.data)) + geom_mosaic(aes(product(smoke, fuel_bin), fill = smoke)) + labs(x= "fuel type", y = "smoking status", title = 'fuel type (clean/solid) used based on smoking status')
```

Smoking status would not be a covariate. Most women do not smoke.

# USING max birthweight of 4500g
##education vs birth weight
```{r}
is.na(gh.data$c_weight) <- which(gh.data$c_weight >= 4501)
ggplot(na.omit(gh.data), aes(x= education, y= c_weight))+geom_boxplot(outlier.shape = NA)+ ylim(0,6000)+ ylab("birth weight")
```

Those with no education or primary education seem to have babies with a lower mean weight.


##birth weight vs wealth
```{r}
ggplot(na.omit(gh.data), aes(x= wealth, y= c_weight))+geom_boxplot(outlier.shape = NA)+ ylim(0,6000)
```

The mean birth weight for each category seem to be lower for those who are poorest or poor.

## birth weight vs residence
```{r}
ggplot(na.omit(gh.data), aes(x= residence, y= c_weight))+geom_boxplot(outlier.shape = NA)+ ylim(0,6000)
```

Average Birth weights for those in rural settings seem to be quite lower than those in the urban setting.


## birth weight vs gender
```{r}
gh.data$gender <- as.factor(gh_data$gender)
ggplot(na.omit(gh.data), aes(x= gender, y= c_weight))+geom_boxplot(outlier.shape = NA)+ ylim(0,6000)+ labs(x= "gender", y= "birth weight")
```

boys seem to have a higher average birth weight than girls

